### 2025-07-08 pca.qmd
20250707 - PCA example

```{r}

```

@@

>>>>

```{r}
library("psych")
library("nFactors")
library("tidyverse")
```

### Load Data {#sec-pcaLoadData}

```{r}
#| eval: false
#| include: false

load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_weekly.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/nfl_nextGenStats_weekly.RData", fsep = ""))
```

```{r}
load(file = "./data/player_stats_weekly.RData")
load(file = "./data/nfl_nextGenStats_weekly.RData")
```

### Prepare Data {#sec-pcaPrepareData}

#### Merge Data {#sec-pcaPrepareDataMerge}

```{r}
dataMerged <- full_join(
  player_stats_weekly,
  nfl_nextGenStats_weekly %>% select(-any_of(c("player_display_name","completions","attempts","receptions","targets"))),
  by = c("player_id" = "player_gsis_id","season","season_type","week"),
)
```

#### Specify Variables {#sec-pcaPrepareDataSpecifyVars}

```{r}
pcaVars <- c(
  "completions","attempts","passing_yards","passing_tds","passing_interceptions",
  "sacks_suffered","sack_yards_lost","sack_fumbles","sack_fumbles_lost",
  "passing_air_yards","passing_yards_after_catch","passing_first_downs",
  "passing_epa","passing_cpoe","passing_2pt_conversions","pacr","pass_40_yds",
  "pass_inc","pass_comp_pct","fumbles","two_pts",
  "avg_time_to_throw","avg_completed_air_yards","avg_intended_air_yards",
  "avg_air_yards_differential","aggressiveness","max_completed_air_distance",
  "avg_air_yards_to_sticks","passer_rating", #,"completion_percentage"
  "expected_completion_percentage","completion_percentage_above_expectation",
  "avg_air_distance","max_air_distance")
```

#### Standardize Variables {#sec-pcaPrepareDataStandardize}

```{r}
dataForPCA <- dataMerged
dataForPCA[pcaVars] <- scale(dataForPCA[pcaVars])
```

@@

## Example of Principal Component Analysis {#sec-pcaExample}

```{r}
#| label: fig-pcaScreePlot1
#| fig-cap: "Scree Plot."
#| fig-alt: "Scree Plot."

psych::fa.parallel(
  x = dataForPCA[pcaVars],
  fa = "pc"
)
```

```{r}
#| label: fig-pcaScreePlot2
#| fig-cap: "Scree Plot."
#| fig-alt: "Scree Plot."

screeDataPCA <- nFactors::nScree(
  x = cor(
    dataForPCA[pcaVars],
    use = "pairwise.complete.obs"),
  model = "components")

nFactors::plotnScree(screeDataPCA)
```

```{r}
psych::vss(
  dataForPCA[pcaVars],
  rotate = "oblimin",
  fm = "pc")
```

```{r}
pca1ComponentOblique <- psych::principal(
  dataForPCA[pcaVars],
  nfactors = 1,
  rotate = "oblimin")

pca2ComponentOblique <- psych::principal(
  dataForPCA[pcaVars],
  nfactors = 2,
  rotate = "oblimin")

pca3ComponentOblique <- psych::principal(
  dataForPCA[pcaVars],
  nfactors = 3,
  rotate = "oblimin")

pca4ComponentOblique <- psych::principal(
  dataForPCA[pcaVars],
  nfactors = 4,
  rotate = "oblimin")

pca5ComponentOblique <- psych::principal(
  dataForPCA[pcaVars],
  nfactors = 5,
  rotate = "oblimin")

pca6ComponentOblique <- psych::principal(
  dataForPCA[pcaVars],
  nfactors = 6,
  rotate = "oblimin")

pca7ComponentOblique <- psych::principal(
  dataForPCA[pcaVars],
  nfactors = 7,
  rotate = "oblimin")

pca8ComponentOblique <- psych::principal(
  dataForPCA[pcaVars],
  nfactors = 8,
  rotate = "oblimin")

pca9ComponentOblique <- psych::principal(
  dataForPCA[pcaVars],
  nfactors = 9,
  rotate = "oblimin")
```

```{r}
pca1ComponentOblique
pca2ComponentOblique
pca3ComponentOblique
pca4ComponentOblique
pca5ComponentOblique
pca6ComponentOblique
pca7ComponentOblique
pca8ComponentOblique
pca9ComponentOblique
```

Here are the variables that had a standardized component loading of 0.4 or greater on each component:

```{r}
component1vars <- c(
  "completions","attempts","passing_yards","passing_tds","passing_interceptions",
  "sacks_suffered","sack_yards_lost","sack_fumbles","sack_fumbles_lost",
  "passing_air_yards","passing_yards_after_catch","passing_first_downs",
  #"passing_epa","passing_cpoe","passing_2pt_conversions","pacr","pass_40_yds",
  "pass_inc","fumbles")#,"two_pts","pass_comp_pct",
  #"avg_time_to_throw","avg_completed_air_yards","avg_intended_air_yards",
  #"avg_air_yards_differential","aggressiveness","max_completed_air_distance",
  #"avg_air_yards_to_sticks","passer_rating", #,"completion_percentage"
  #"expected_completion_percentage","completion_percentage_above_expectation",
  #"avg_air_distance","max_air_distance")

component2vars <- c(
  #"completions","attempts","passing_yards","passing_tds","passing_interceptions",
  #"sacks_suffered","sack_yards_lost","sack_fumbles","sack_fumbles_lost",
  "passing_air_yards",#"passing_yards_after_catch","passing_first_downs",
  "pacr",#"passing_epa","passing_cpoe","passing_2pt_conversions","pass_40_yds",
  #"pass_inc","pass_comp_pct","fumbles","two_pts",
  "avg_completed_air_yards","avg_intended_air_yards",#"avg_time_to_throw",
  "aggressiveness","max_completed_air_distance",#"avg_air_yards_differential",
  "avg_air_yards_to_sticks",#"passer_rating", #,"completion_percentage"
  "expected_completion_percentage",#"completion_percentage_above_expectation",
  "avg_air_distance","max_air_distance")

component3vars <- c(
  "passing_tds",#"completions","attempts","passing_yards","passing_interceptions",
  #"sacks_suffered","sack_yards_lost","sack_fumbles","sack_fumbles_lost",
  #"passing_air_yards","passing_yards_after_catch","passing_first_downs",
  "passing_epa","passing_cpoe","pass_40_yds",#"passing_2pt_conversions","pacr",
  "pass_comp_pct",#"fumbles","two_pts","pass_inc",
  "avg_air_yards_differential","max_completed_air_distance",
  "avg_time_to_throw","avg_completed_air_yards","avg_intended_air_yards",
  #"aggressiveness",
  "passer_rating",#"avg_air_yards_to_sticks", #,"completion_percentage"
  "completion_percentage_above_expectation")#,"expected_completion_percentage",
  #"avg_air_distance","max_air_distance")
```

```{r}
pca3ComponentOblique$scores %>% 
  na.omit() %>% 
  head()
```

```{r}
pca3ComponentOblique$scores[which.max(pca3ComponentOblique$scores[,1]),]
dataMerged[which.max(pca3ComponentOblique$scores[,1]), c("player_display_name", "season", "week", component1vars)]

pca3ComponentOblique$scores[which.min(pca3ComponentOblique$scores[,1]),]
dataMerged[which.min(pca3ComponentOblique$scores[,1]), c("player_display_name", "season", "week", component1vars)]
```

```{r}
pca3ComponentOblique$scores[which.max(pca3ComponentOblique$scores[,2]),]
dataMerged[which.max(pca3ComponentOblique$scores[,2]), c("player_display_name", "season", "week", component2vars)]

pca3ComponentOblique$scores[which.min(pca3ComponentOblique$scores[,2]),]
dataMerged[which.min(pca3ComponentOblique$scores[,2]), c("player_display_name", "season", "week", component2vars)]
```

```{r}
pca3ComponentOblique$scores[which.max(pca3ComponentOblique$scores[,3]),]
dataMerged[which.max(pca3ComponentOblique$scores[,3]), c("player_display_name", "season", "week", component3vars)]

pca3ComponentOblique$scores[which.min(pca3ComponentOblique$scores[,3]),]
dataMerged[which.min(pca3ComponentOblique$scores[,3]), c("player_display_name", "season", "week", component3vars)]
```
